---
title: "README"
author: "A Gaba"
date: "April 21, 2015"
output: html_document
---
```{r}
library(reshape2)
library(dplyr)
```
# Data sets

There are three major datasets that we create, respectively, measurements, observations and subjects. We create therefore one function for retrieving and formating the data for each of these data sets. 

## Observations table
The following function read the training and test data sets and binds the resulting tables by row into one table.
Then, a subset of the observations table is extracted including only mean and stddev observations.
Finally, we read the features table to assign titles properly to the table of observations.
```{r}
get_obs_table <- function() {
                            # read features
                            features = read.table("UCI\ HAR\ Dataset/features.txt", header=FALSE, sep="")
                            # read test data
                            test_tab = read.table("UCI\ HAR\ Dataset/test/X_test.txt", header=FALSE, sep="")
                            dim(test_tab)
                            # read train data
                            train_tab = read.table("UCI\ HAR\ Dataset/train/X_train.txt", header=FALSE, sep="")
                            dim(train_tab)                            
                            # merge train and test data
                            obs_table = rbind(test_tab, train_tab)                            
                            # extract only the measurements concerning the mean and standard deviation for each measurement.
                            indexes = which(grepl( "*mean\\(\\)*|*std\\(\\)*", features$V2))
                            sub_obs_table = select(obs_table, indexes)                            
                            # get the descriptions of the measurements                            
                            descriptions = features[which(grepl( "*mean\\(\\)*|*std\\(\\)*", features$V2)),][2]
                            names(sub_obs_table) = descriptions$V2
                            return(sub_obs_table)            
}
```

## Subjects table
Similarly, we put together training and test data and assign a title to the resulting variable. 
```{r}
get_subjects_table <- function() {
                            # read test data
                            test_subj_tab = read.table("UCI\ HAR\ Dataset/test/subject_test.txt", header=FALSE, sep="")
                            dim(test_subj_tab)
                            # read train data
                            train_subj_tab = read.table("UCI\ HAR\ Dataset/train/subject_train.txt", header=FALSE, sep="")
                            dim(train_subj_tab)
                            # merge train and test data 
                            subjects_table = rbind(test_subj_tab, train_subj_tab)
                            # name the subject table
                            names(subjects_table) = c("SubjectID")
                            return(subjects_table)
}
```

## Activities table
After putting together the training and test data sets, we retrieve the table of activity lables and join the two resulting tables.
```{r}
get_activities_table <- function() {                            
                            # read test data
                            test_activity_tab = read.table("UCI\ HAR\ Dataset/test/y_test.txt", header=FALSE, sep="")
                            dim(test_activity_tab)
                            # read train data
                            train_activity_tab = read.table("UCI\ HAR\ Dataset/train/y_train.txt", header=FALSE, sep="")
                            dim(train_activity_tab)    
                            # merge train and test data
                            activities_table = rbind(test_activity_tab, train_activity_tab)
                            # read the activity labels
                            activity_labels = read.table("UCI\ HAR\ Dataset/activity_labels.txt", header=FALSE, sep="")
                            # join activity_labels with the activity table to assign a label to each activity code
                            complete_activity = merge(activities_table, activity_labels, by="V1")
                            # name the columns
                            names(complete_activity) = c("ActivityCode", "ActivityDescription")
                            return(complete_activity)
}
```

## Finalizing steps
After retrieving the three tables, we bind them together by column and form a big *wide* table. As wide tables are not flexible for performing manipulations, we transform it into a long table by using the function *melt*.
```{r}
observations_table = get_obs_table()
activities_table = get_activities_table()
subjects_table = get_subjects_table()

# bind by column the three tables
complete_table = cbind(subjects_table, activities_table, observations_table)
# head(complete_table)

# reshape to transform the table into a long one
long_table <- melt(complete_table, id.vars=c("SubjectID", "ActivityCode","ActivityDescription"), variable.name="Measurement")
head(long_table)
```

In order to compute the mean of the measurements for each subject, we use the library *dplyr* as follows.
```{r}
# grouping the observations by subject, activity and measurement
grouped = group_by(long_table, SubjectID, ActivityCode, ActivityDescription, Measurement)
head(grouped)

# compute the mean of the each measurement
final = summarise(grouped, meanVal=mean(value))
head(final)
```
Finally, we write the result in a file.
```{r}
# write to file
write.table(final, file="tidy_data.txt", row.names=FALSE)
print(final)
```
